Paper 1: Instrument development

### 2024-09-03 Rmd/Paper1.Rmd
First complete version of Rmd for paper 1.

library(readODS)                            # read .ods-files. Kilde: https://github.com/ropensci/readODS/
library(readxl)                             # read .xlsx-files. Kilde: https://readxl.tidyverse.org/

library(pander)                             # print pretty tables
library(dplyr)                              # rename columns
library(tidyr)                              # pivot_longer

library(ggplot2)

# multicollinearity checks
library("corrplot")                         # multicollinearity 
library(car)                                # vif       
library("olsrr")                            # vif

library(psych)                              # for skew
library(mice)                               # for missing

@@

@@

```{r select_data_for_screening}
# emne2_anon (uten post)
data_TPACK_pilot_minuspost <- data_TPACK_pilot[
  data_TPACK_pilot$datasett != "post" & 
  data_TPACK_pilot$datasett != "pilot_trelis_q",]

@@

## Step 5: Item reduction

@@

@@

@@

```{r remove_NMAR_TRELISQ}
# remove NMAR + NTNU (19)
data_TPACK_PST_noNMAR <-
  data_trelis_q_ren[-c(19,34,44,45,46,95,102),]

@@

@@

@@

@@

>>>>

### read files
library(readODS)                            # read .ods-files. Kilde: https://github.com/ropensci/readODS/
library(readxl)                             # read .xlsx-files. Kilde: https://readxl.tidyverse.org/

### presentation
library(pander)                             # print pretty tables
library(dplyr)                              # rename columns
library(tidyr)                              # pivot_longer
library(ggplot2)

### screening
library("corrplot")                         # multicollinearity
library(car)                                # vif       
library("olsrr")                            # vif
library(mice)                               # for missing

### EFA
library(psych)                              # for skew and EFA

### CFA
library(lavaan)
library(semTools)                           # to find omega

@@

```{r}
# clean environment to reset for next step
# ref.: https://www.statology.org/clear-environment-in-r/ [3.9.2024]
rm(list=ls())
```

## Step 5: Item reduction (assumption checking)

```{r}
do_screening <- 1 # skip screening during testing
```

```{r}
# load data
load(file="../data/processed_data/tpack-pilot_2022.Rdata") 
load(file="../data/processed_data/tpack-trelis_q_paper1_2023.Rdata")
```

@@

```{r select_data_for_screening_step5}
# emne2_anon (uten post)
data_TPACK_pilot_minuspost <- data_TPACK_pilot[
  data_TPACK_pilot$datasett != "post" & 
  data_TPACK_pilot$datasett != "pilot_trelis_q",]

@@

@@

```{r}
# clean environment to reset for next step
# ref.: https://www.statology.org/clear-environment-in-r/ [3.9.2024]
rm(list=ls())
```

@@

Here I do an exploratory factor analysis (EFA) to examine how the items correlate and can be structured. The aim is to reduce the items to a limited set of factors that represent the data.

```{r}
# load data
load(file="../data/processed_data/tpack-pilot_2022.Rdata") 
```

```{r items_EFA}
TK_pos <- c("TK1","TK2","TK3","TK4","TK5","TK6","TK7","TK8")
PK_pos <- c("PK1","PK2","PK3","PK4","PK5","PK6")
CK_pos <- c("CK1","CK2","CK3","CK4","CK5")
PCK_pos <- c("PCK1","PCK2","PCK3","PCK4","PCK5")
TCK_pos <- c("TCK1","TCK2","TCK3","TCK4")
TPK_pos <- c("TPK1","TPK2","TPK3","TPK4","TPK5")
TPACK_pos <- c("TPACK1","TPACK2","TPACK3","TPACK4","TPACK5")

```{r select_data_for_screening_step6}
# emne2_anon (uten post)
data_TPACK_pilot_minuspost <- data_TPACK_pilot[
  data_TPACK_pilot$datasett != "post" & 
  data_TPACK_pilot$datasett != "pilot_trelis_q",]
```

```{r remove_collinear_items_EFA}
### pilot
data_TPACK_pilot_minuspost <- data_TPACK_pilot_minuspost %>%
    # 25.5.2023: velg bort items som ikke er i TRELIS-Q
    select(-c("TK8","PK2","PK4","PK5","PK6","PCK1","PCK3","PCK4","TCK2","TCK3",
              "TPK2","TPK3","TPK5","TPACK1","TPACK5",
              "TPACK3", # colinear in trelis-q; "TK1","TPK1",
              "TK6","TK4", "PK1", "PK3"))

### Describe variables

To get an overview of the quality of the data and to see what data it is appropiate to use further in the analysis.

First count distribution of answers on the likert-scale used.

```{r likert_distribution}
# kilde: https://solomonkurz.netlify.app/post/2021-05-11-yes-you-can-fit-an-exploratory-factor-analysis-with-lavaan/
data_TPACK_pilot_minuspost[,colnames(data_TPACK_pilot_minuspost) %in% unlist(pilot_list)] %>% 
  pivot_longer(everything(),values_to="value") %>% 
  count(value) %>% 
  mutate(percent = (100 * n / sum(n)) %>% round(digits = 1))
```

Plot histogram for each items, to see how they compare.

```{r histogram}
# plot multi-panel histograms with facet_wrap
# kilde: https://www.zevross.com/blog/2019/04/02/easy-multi-panel-plots-in-r-using-facet_wrap-and-facet_grid-from-ggplot2/

data_long <- pivot_longer(data_TPACK_pilot_minuspost, 
                          cols=colnames(data_TPACK_pilot_minuspost)[colnames(data_TPACK_pilot_minuspost) %in% unlist(pilot_list)],
                          names_to="item", 
                          values_to="likert_verdi",
                          values_drop_na = TRUE)

ggplot(data_long, aes(factor(likert_verdi))) +
  geom_histogram(stat = "count") + 
  facet_wrap("item")
```

<!-- Poor items: y10 (CK2), y13 (CK5) (because of range=2). -->

### Preliminary analysis

"It is recommended to carry out a preliminary analysis of the metric quality of the items to subject the most adequate items to EFA. For this purpose, it is recommended to analyze and report the mean, standard deviation, and item-test correlation of each one of the items, as well as the Cronbach’s alpha of the scales of the test. The researcher should decide whether to eliminate certain items" (Izquierdo et al, 2014, p. ??)

```{r help_function_correlation_matrix}
### HELP FUNCTION

antall_desimaler <- 2

# print effect size with p-value
print_starvalue <- function (pvalue, effectsize, bonferroni=1) {
  ifelse (pvalue < 0.001/bonferroni, star <- "***",
  ifelse (pvalue < 0.01/bonferroni, star <- "**",
  ifelse (pvalue < 0.05/bonferroni, star <- "*",
          star <- "")))
  return (paste(format(effectsize,
                       digits=antall_desimaler,
                       nsmall=antall_desimaler),
                star, sep="")) # antall desimaler=2
}

create_matrix <- function (df, method="pearson") {
  # create correlation matrix for df (and covariance matrix combined, cf. Weston, 2006, table 2, p. 736) (cov lower left, cor upper right)

  # since ordered data, use spearman; option in function call?

  # left side
  # use casewise deletion if missing
  cov_matrix     <- cov(df, use="complete.obs", method=method)
  # right side
  cor_matrix     <- data.frame(matrix(ncol = ncol(df), nrow = 0)) # correlation matrix

  i              <- 1          # start posisjon
  teller         <- 0          # hvor mange plot lager jeg?
  for (var1 in df[1:ncol(df)-1]) {
    i            <- i + 1

    # create right side: correlation 
    right_side   <- data.frame(matrix(ncol = 0, nrow = 1)) # row values
    for (var2 in df[i:ncol(df)]) {
      teller     <- teller + 1   
      cor_var1_var2 <- cor.test(var1,var2, method=method)
      est_p      <- print_starvalue(cor_var1_var2$p.value,
                                    cor_var1_var2$estimate)
      right_side <- cbind(right_side,est_p) # add column
    }

    # add left-side of matrix: covariance
    left_side    <- format(cov_matrix[i-1,1:i-1],
                           digits=antall_desimaler)

    temp_row     <- c(left_side, right_side)  # merge left and right
    names(temp_row) <- colnames(df)
    cor_matrix   <- rbind(cor_matrix, temp_row)   # add row
  }

  # add final row
  final_row      <- format(cov_matrix[i,1:i],digits=antall_desimaler)
  names(final_row) <- colnames(df)
  cor_matrix     <- rbind(cor_matrix, final_row)

  # add row with mean and sd based on rows/persons
  row_mean       <- round(colMeans(df, na.rm=TRUE),
                          digits=antall_desimaler)
  row_sd         <- format(psych::SD(df),
                          digits=antall_desimaler)
  cor_matrix     <- rbind(cor_matrix,"mean"=row_mean)
  cor_matrix     <- rbind(cor_matrix,"sd"=row_sd)

  # set names of columns and rows
  row.names(cor_matrix) <- c(colnames(df),"mean","sd")
  return(cor_matrix)
}

```{r calculate_mean_sd}
# select Likert columns
data_pilot <- data_TPACK_pilot_minuspost[,colnames(data_TPACK_pilot_minuspost) %in% unlist(pilot_list)]

# remove items with over 50 % NA
df_noNA <- data_pilot[ , colSums(is.na(data_pilot)) < nrow(data_pilot)/2] 
```

#### eigenvalues

```{r}
# check for collinearity; values near or below 0
ev <- eigen(cor(df_noNA[,colnames(df_noNA) %in%
                                unlist(pilot_list)],
                use="pairwise.complete.obs"), only.values=TRUE)

lag_eigenvalue_tabell (ev$values)
```

#### item-test correlation

Check to see if an item corresponds to the average test-result for the intended factor (e.g., if the TK-items correspond to the other TK-items).

Ref.: https://www.questionmark.com/resources/blog/psychometrics-101-item-total-correlation/ https://en.wikipedia.org/wiki/Item-total_correlation

```{r item_test}
# get data for each test / construct
TK    <- df_noNA[,colnames(df_noNA) %in% TK_pos]
PK    <- df_noNA[,colnames(df_noNA) %in%PK_pos]
CK    <- df_noNA[,colnames(df_noNA) %in%CK_pos] # NA
TCK   <- df_noNA[,colnames(df_noNA) %in%TCK_pos]
TPK   <- df_noNA[,colnames(df_noNA) %in%TPK_pos]
PCK   <- df_noNA[,colnames(df_noNA) %in%PCK_pos]
TPACK <- df_noNA[,colnames(df_noNA) %in% TPACK_pos]

# check correlation of items to construct
pander(create_matrix(data.frame(TK,    "TK"=rowMeans(TK, na.rm=TRUE))))
pander(create_matrix(data.frame(PK,    "PK"=rowMeans(PK, na.rm=TRUE))))
pander(create_matrix(data.frame(CK,    "CK"=rowMeans(CK, na.rm=TRUE))))
pander(create_matrix(data.frame(TCK,   "TCK"=rowMeans(TCK, na.rm=TRUE))))
pander(create_matrix(data.frame(TPK,   "TPK"=rowMeans(TPK, na.rm=TRUE))))
pander(create_matrix(data.frame(PCK,   "PCK"=rowMeans(PCK, na.rm=TRUE))))
pander(create_matrix(data.frame(TPACK, "TPACK"=rowMeans(TPACK, na.rm=TRUE))))

TCK2 has low item-test correlation.

#### Reliability: Cronbach's alpha

Assumptions for calculating Cronbach's alpha: uni-dimensionality (one factor = factor analysis) and tau-equality (equal covariance) [@revelle_reliability_2019].

##### TK

```{r uni-dimensionality}
parallel <- fa.parallel(TK,fm='uls',fa='fa',SMC=TRUE) #, cor="poly", correct=0 ,use="complete.obs") #)
parallel$fa.values
```

```{r tau-equality, warning=T}
pander(create_matrix(TK))
```

Forutsetninger ikke oppfylt: ikke uni-dimensional (2 faktorer fra parallell analysis, og 1 fra eigenvalue), og ikke tau-equality (TK8 passer ikke på tau-equality).

```{r alpha}
psych::alpha(TK)
```

##### CK

```{r warning=FALSE}
parallel <- fa.parallel(CK,fm='uls',fa='fa',SMC=TRUE,  cor="poly", correct=0,use="complete.obs") #,cor="poly")
parallel$fa.values
pander(create_matrix(CK))
psych::alpha(CK) 
```

Uni-dimensional, og nesten tau-equality. CK4 "ødelegger". CK4 er allerede tagget for fjerning.

##### PK

```{r warning=FALSE,eval=F}
parallel <- fa.parallel(PK,fm='uls',fa='fa',SMC=TRUE,  cor="poly", correct=0,use="complete.obs") #,cor="poly")
parallel$fa.values
pander(create_matrix(PK))
psych::alpha(PK)

Uni-dimensional. Bortimo tau-equality. PK4 og PK5 "ødelegger" litt. Begge disse er allerede tagget for fjerning. PK6 senker også reliabiliteten. (Denne er dog tatt med videre til TRELIS-q)

##### TCK

```{r warning=FALSE}
parallel <- fa.parallel(TCK,fm='uls',fa='fa',SMC=TRUE,  cor="poly", correct=0,use="complete.obs") #,cor="poly")
parallel$fa.values
pander(create_matrix(TCK))
psych::alpha(TCK)

Uni-dimensional. Ikke tau-equality. TCK1 ødelegger.

##### TPK

```{r warning=FALSE,eval=T}
parallel <- fa.parallel(TPK,fm='uls',fa='fa',SMC=TRUE) #,  cor="poly", correct=0,use="complete.obs") #,cor="poly")
parallel$fa.values
pander(create_matrix(TPK))
psych::alpha(TPK)

Uni-dimensional og omtrent tau-equality.

##### PCK

```{r warning=FALSE}
parallel <- fa.parallel(PCK,fm='uls',fa='fa',SMC=TRUE,  cor="poly", correct=0,use="complete.obs") #,cor="poly")
parallel$fa.values
pander(create_matrix(PCK))
psych::alpha(PCK)

Uni-dimensional og omtrent tau-equality. PCK1 og PCK3 kan fjernes. (men PCK1 er tatt med videre)

##### TPACK

```{r warning=FALSE, eval=T}
  parallel <- fa.parallel(TPACK,fm='uls',fa='fa',SMC=TRUE,
                          cor="poly", correct=0,use="complete.obs") #,cor="poly")
  parallel$fa.values
  pander(create_matrix(TPACK))
  psych::alpha(TPACK)

Uni-dimensional og halvveis tau-equality. TPACK1 "ødelegger" litt.

#### Results of preliminary analysis

TK7 and CK4 stand apart as poor items.

Reset data for EFA

```{r}
# remove columns
data_modified <- within(data_TPACK_pilot_minuspost, rm("TK7","CK4"))

# pick Likert columns
df <- data_modified[,colnames(data_modified) %in% unlist(pilot_list)]

# remove respondents with more than 50% NA
df_run1 <- df[ , colSums(is.na(df)) < nrow(df)/2]

### EFA

The aim of the exploratory factor analysis is to find the lowest number of items that is needed to represent the dataset.

Estimation is done using maximum-likelihood estimation. Rotation is oblique (oblimin). tenBerge is used to estimate scores (ref. Psych-package help-file).

#### run 1

The suitability of the data for doing an EFA is tested by calculating KMO og Bartlett.

```{r}
# KMO og Bartlett
KMO_base      <- KMO(df_run1) # addition of PCK1 changed KMO from 0.5 to 0.36
bartlett_base <- cortest.bartlett(df_run1, n=5,diag=TRUE)
```

The number of factors is estimated using parallel analysis [@worthington_scale_2006].

```{r parallel_analysis_run1}
# number of factors: eigenvalue, scree or parallel (or theoretical)
# polychoric is used for ordered data with skewness, Watkins (2018)

parallel_base <- fa.parallel(df_run1,fm='mle',fa='fa',SMC=TRUE,  use="complete.obs") #, )correct=0,cor="poly",

#parallel <- fa.parallel(df,fm='alpha',fa='fa',nfactors=7,use="complete.obs",cor='poly', error.bars=TRUE, SMC=TRUE) # complete.obs = delete NA; poly for fa recommended av Revelle & Condon, 2019
```

```{r}
# eigenvalues
lag_eigenvalue_tabell (parallel_base$fa.values)

print(format(list("KMO" = KMO_base$MSA, "Bartlett p =" = bartlett_base$p.value),digits=2))
```

Due to the ambiguity of the parallel analysis, several factor solutions are tested (i.e., 2-factor, 3-factor and 4-factor).

##### 2-factor

```{r efa_ordinal2}
# FA
# ref.: https://www.rdocumentation.org/packages/psych/versions/2.2.5/topics/fa

# print covariation matrix all items
# bruk fm=uls, ikke pa (Flora et al, 2012)
# bruk polykorisk korrelasjon (Flora et al, 2012)
# use covariance matrix (correlation matrix fails because of missing)

model_2factor_run1 <- fa(df_run1, nfactors = 2,rotate = "oblimin", fm="mle", scores="tenBerge") # missing=TRUE,impute="mean", ,oblique.scores=TRUE, cor="poly", correct=0 covar=TRUE,

print.psych(fa.sort(model_2factor_run1))
```

##### 3-factor

```{r efa_ordinal3}
# FA
# ref.: https://www.rdocumentation.org/packages/psych/versions/2.2.5/topics/fa

# print covariation matrix all items
# bruk fm=uls, ikke pa (Flora et al, 2012)
# bruk polykorisk korrelasjon (Flora et al, 2012)
# use covariance matrix (correlation matrix fails because of missing)

model_3factor_run1 <- fa(df_run1, nfactors = 3,rotate = "oblimin", fm="mle", scores="tenBerge") # missing=TRUE,impute="mean", covar=TRUE,,oblique.scores=TRUE, cor="poly", correct=0

print.psych(fa.sort(model_3factor_run1))
```

##### 4-factor

```{r efa_ordinal4}
# FA
# ref.: https://www.rdocumentation.org/packages/psych/versions/2.2.5/topics/fa

# print covariation matrix all items
# bruk fm=uls, ikke pa (Flora et al, 2012)
# bruk polykorisk korrelasjon (Flora et al, 2012)
# use covariance matrix (correlation matrix fails because of missing)

model_4factor_run1 <- fa(df_run1, nfactors = 4,rotate = "oblimin", fm="mle", scores="tenBerge") # missing=TRUE,impute="mean", covar=TRUE,,oblique.scores=TRUE, cor="poly", correct=0

print.psych(fa.sort(model_4factor_run1))
```

```{r}
# eigenvalues
print(model_3factor_run1$values)
```

Eigenvalues suggests 3 factors

The cutoff for significant loadings ...

```{r}
# cutoff verdi gitt av formel i Watkins, 2018 (avsnitt: Interpretation of results)
n <- model_3factor_run1$n.obs
c <- 4.084/sqrt(n-2)   # n=100 gir 3.92 p<5% (z-dist); n=30 gir 4.084 for p<5% (t-dist)
print(c)
# bare print loadings med verdier over 0.32
print(model_3factor_run1$loadings,cutoff = c)

print.psych(model_3factor_run1)
```

Weak communality ...

```{r communality}
# h2 = communality; bare print items med communality under 0.4 siden disse kan fjernes (Worthington, 2006); eller 0.6 for små utvalg? (ref)
# kilde: https://stackoverflow.com/questions/30467834/r-print-cutoff-values-under-a-certain-value
print(model_3factor_run1$communality[model_3factor_run1$communality<0.6])
```

Remove poor items from run 1 due to cross-loadings ...

```{r}
# 26.5.2023, remove cross-loadings
df_run2 <- within(df_run1, rm("TK5"))
```

Kan vurdere å beholde CK1 med h2 = 0.37.

#### run 2

The EFA is repeated, since an item was removed.

Suitability of the data for doing an EFA.

```{r}
# KMO og Bartlett
KMO_base      <- KMO(df_run2) # addition of PCK1 changed KMO from 0.5 to 0.36
bartlett_base <- cortest.bartlett(df_run2, n=4,diag=TRUE)
```

```{r parallel_analysis_run2}
# fm = mle, alpha, pa

#if (do_parallel == 1) {

# number of factors: eigenvalue, scree or parallel (or theoretical)
# parallel analysis (Worthington, 2006)
# polychoric is used for ordered data with skewness, Watkins (2018)
  # "pa (parallel analysis) most accurately estimated the number of factors" (Li et al., 2020)

parallel_base <- fa.parallel(df_run2,fm='mle',fa='fa',SMC=TRUE) #, ),  correct=0,cor="poly",use="complete.obs"
#parallel <- fa.parallel(df,fm='alpha',fa='fa',nfactors=7,use="complete.obs",cor='poly', error.bars=TRUE, SMC=TRUE) # complete.obs = delete NA; poly for fa recommended av Revelle & Condon, 2019
```

```{r}
# eigenvalues
#parallel$fa.values
lag_eigenvalue_tabell (parallel_base$fa.values)

print(format(list("KMO" = KMO_base$MSA, "Bartlett p =" = bartlett_base$p.value),digits=2))
```

##### 2 factor

```{r}
model_2factor_run2 <- fa(df_run2, nfactors = 2, rotate = "oblimin", fm="mle", scores="tenBerge") # ,oblique.scores=TRUE, cor="poly", correct=0

print.psych(fa.sort(model_2factor_run2))
```

##### 3 factor

```{r}
# equamax for complex structure, ref. Schmitt & Sass, 2011
# CI with bootstrapping, n.iter=1000
# p = .8 (Cohen, 1990)
model_3factor_run2 <- fa(df_run2, nfactors = 3, rotate = "oblimin", fm="mle", scores="tenBerge") #, missing=TRUE, impute="mean") # n.iter=1000, p=.8,  #,oblique.scores=TRUE, cor="poly", correct=0)

print.psych(fa.sort(model_3factor_run2))

#factor_model2
```

##### 4 factor

```{r}
model_4factor_run2 <- fa(df_run2, nfactors = 4, rotate = "oblimin", fm="mle", scores="tenBerge") #,oblique.scores=TRUE, cor="poly", correct=0)

print.psych(fa.sort(model_4factor_run2))
```

Check 2nd order solution.

```{r}
# korrelation matrix factors
#format(round(factor_model$r.scores,1), nsmall=1) 
order2 <- model_3factor_run2$Phi # 2nd order factors
print(order2)
```

```{r}
# eigenvalues
print(model_3factor_run2$values)

# loadings
print.psych(model_3factor_run2) #, cut=c)
```

Remove items "with factor loadings less than .32 or cross-loadings less than .15 difference from an item’s highest factor loading. In addition, they should also delete items that contain absolute loadings higher than a certain value (e.g., .32) on two or more factors." (Worthington and Whittaker, 2006).

#### Assessment

Felles varians: hvor mye felles varians forklarer faktorene? Faktorene bør tilsammen forklare minst 50% av variansen (Christophersen, 2018).

```{r}
p <- model_3factor_run2$Vaccounted
print(p)                           # se på cumulative var
```

Kommunaliteten (h2) beskriver hvor stor del av variansen til indikatoren som faktorene "forklarer". Residualvariansen (u2) er den uforklarte variansen til indikatoren (item/spørsmålet). Det er i praksis vanskelig å få særlig høyere kommunalitet enn 65% i følge Christophersen (2018).

```{r communality7}
# Mundfrom et al., 2005: Råd: høy kommunalitet (>0.6) 
# dette er unstandardized loadings / covariance matrix ??
print(model_3factor_run2$communality[model_3factor_run2$communality>0.6])
# dette er standardized loadings ??
#print(efa7_ml_pro$communalities[efa7_ml_pro$communalities>0.6])

# avoid below 0.4
print(model_3factor_run2$communality[model_3factor_run2$communality<0.4])
```

Vurder modellen: Hvor sterkt korrelerer indikatorene i modellen? Indikatorer med r>0.80 er for sterkt korrelert (Christophersen, 2018). Indikatorer som korrelerer svært svakt kan tyde på at et vesentlig begrepsaspekt er utelatt av modellen.

Når det er korrelerte faktorer vi bruker, så vil strukturmatrisen gi korrelasjonene, mens pattern matrisen gir loadings.

```{r}
# for sterke korrelasjoner
print(model_3factor_run2$Structure, cutoff=0.8)

model_3factor_run2$Structure

# meget svake korrelasjoner
#efa7_ml_pro$Structure[efa7_ml_pro$Structure<0.1]
```

restledd skal helst ikke være større enn +- 0.05

```{r}
round(model_3factor_run2$residual, digits=2)
```

```{r}
# check reliability
#cronbach.alpha(df_run2, na.rm=TRUE)

# tegn factor_model
stimodel <- fa.diagram(model_3factor_run2)
#print(stimodel)
```

```{r}
# compare with intervall pearson (appx normal distributed)
pander(create_matrix(df_run2))
```

### McDonald's omega

Fra help-file til omega: "A recommendation that should be heeded, regardless of the method chosen to estimate \omega_hω, is to always examine the pattern of the estimated general factor loadings prior to estimating \omega_hω. Such an examination constitutes an informal test of the assumption that there is a latent variable common to all of the scale's indicators that can be conducted even in the context of EFA. If the loadings were salient for only a relatively small subset of the indicators, this would suggest that there is no true general factor underlying the covariance matrix. Just such an informal assumption test would have afforded a great deal of protection against the possibility of misinterpreting the misleading \omega_hω estimates occasionally produced in the simulations reported here." (Zinbarg et al., 2006, p 137)"

```{r eval=T}
alpha(df_run2[,c("TK1", "TK2", "TK3")])
alpha(df_run2[,c("CK1", "PCK2", "PCK5", "CK5", "CK3", "CK2")])
alpha(df_run2[,c("TPK1","TPK4","TCK1","TCK4", "TPACK2","TPACK4")]) # , "TPK4"
```

```{r}
omega(df_run2, fm="mle") #, rotate="Promax")
```

```{r}
# clean environment to reset for next step
# ref.: https://www.statology.org/clear-environment-in-r/ [3.9.2024]
rm(list=ls())
```

@@

The following choices were made when determining how to test the data:

* treat the data as interval data.
* missing data: use maximum-likelihood [@mcneish_challenging_2017; @schafer_missing_2002].
* Normal or nonnormal distribution of residuals? Normality is an important assumption for doing test statistics. Nonnormal means skew>2 and kurtosis>9 (ref?). Some of the residuals showed high skewness (i.e., non-normality) (e.g., TPACK2). In such cases, WLS can be an option to OLS [@christophersen_introduksjon_2018]. Some items also showed evidence of being homoscedastic (i.e., non-equal variance in the residuals): CK123, PCK5, TCK4. In case of nonnormal data the Satorra-Bentler scaled chi-square test statistic and robust SE can be used [@flora_old_2012; @mcneish_challenging_2017].
* I aim for approximate fit, not exact fit [@flora_old_2012; weston_brief_2006].

yuan.bentler forutsetter adf-estimering. (McNeish, 2017) som krever at utvalgsstørrelser er minst p(p+1)/2, der p er antall items i modellen. Med p=13 blir det 91 respondenter. Mitt datautvalg er større enn dette (n=103).

test="yuan.bentler(.mplus)" (McNiesh, 2017: small samples? NB: adf)  
se="robust.sem" # standard error  
information="observed" # ref. Schafer & Graham, 2002, p. 162

### Specification of models to be tested

The purpose is to confirm the measurement model, that is, the indicators loading to the latent factors.

The two models are based on the EFA and are nested (e.g., the two_factor model is a constrained version of the three_factor model, i.e., TK and TPACK are merged).

```{r specify_models}
three_factor <- '
PCK   =~ CK123 + PCK2 + PCK5 + CK5 
TK    =~ TK1b + TK2 + TK34 
TPACK =~ TPACK2 + TPACK4 + TCK4 + TPK4 + TCK1 + TPK1
'

two_factor   <- '
PCK   =~ CK123 + PCK2 + PCK5 + CK5
TPACK =~ TK1b + TK2 + TK34 + TPACK2 + TPACK4 + TCK4 + TPK4 + TCK1 + TPK1
'
```

### Test of models

Two models are compared. A two factor and a three factor solution derived from the EFA.

```{r}
# load data
load(file="../data/processed_data/tpack-trelis_q_paper1_2023.Rdata")
```

NMAR are removed from the dataset.

```{r remove_NMAR_TRELISQ}
# remove NMAR + NTNU (19)
CFA_TPACK_PST <- data_TRELIS_Q[-c(19,34,44,45,46,95,102),] 
```

Some help functions are defined.

```{r help_functions_cfa}

### HELP FUNCTIONS
do_cfa        <- function (cfaspec, cfadata) {
  # estimator og missing: Rhemtulla et al. (2012)
  # ref.: https://statistics.ohlsen-web.de/multiple-imputation-with-mice/
  TPACK_PST_0 <- cfa(cfaspec, data= cfadata, std.lv=TRUE,
                     missing="fiml", #robust.two.stage",
                     estimator="ML", #missing="pairwise", 
                     #test="satorra.bentler", 
                     test="yuan.bentler",
                     information="observed",
                     se="robust.huber.white") # robust.sem

  return (TPACK_PST_0)
}

print_model <- function (cfamodel) {
  # print results of CFA

  print("summary")
  print(summary(cfamodel, standardized=TRUE, fit.measures=TRUE))

  print("estimates")
  # CI suggested by Cheung et al. (2023) for discriminant validity < 1 (one-tailed)
  print(parameterEstimates(cfamodel, rsquare = TRUE, level=0.90))

  print("modindices")
  print(modindices(cfamodel, sort=TRUE, maximum.number = 5))            # mi should be below 3.84

  print("reliability")
  rel <- compRelSEM(cfamodel, dropSingle = FALSE)
  print(rel)
  return (rel)   # for later comparison of reliability between models
}

@@

#### two_factor

```{r estimate_2factor}
# egenfaktor
model_two_factor <- do_cfa(two_factor, CFA_TPACK_PST)
rel_two_factor   <- print_model(model_two_factor)
```

#### three_factor

```{r estimate_3factor}
# egenfaktor
model_three_factor <- do_cfa(three_factor, CFA_TPACK_PST)
rel_three_factor   <- print_model(model_three_factor)
```

### Summary of fit

```{r CFA_results_comparison}
# fit indices
columnnames  <- c("AIC", "chi-square", "df", "p-value",
                  "RMSEA", "RMSEA (90%) CI lower","RMSEA (90%) CI upper",
                  "CFI","TLI", "SRMR") #, "chi2/df", 
                 #"chi-square scaled Swain", "p-value Swain", 
                 #"swain chi2/df")

# fit criteria
fit_criteria <- c("x", "x", "x", ">0.05", 
                  "<0.06(0.05-0.10)", "x", "x",
                  ">0.95(0.9-0.95)", ">0.95", "<0.08(0.08-0.15)") #, "<2", "x", ">0.05", "<2") # Weston and Gore (2006); Gallagher and Brown (2013)

fit_2factor  <- round(fitMeasures(model_two_factor,
                  c("aic","chisq.scaled", "df.scaled",
                    "pvalue.scaled",
                    "rmsea.robust", "rmsea.ci.lower.robust", "rmsea.ci.upper.robust",
                    "cfi.robust","tli.robust","srmr")),3)

fit_3factor  <- round(fitMeasures(model_three_factor,
                  c("aic","chisq.scaled", "df.scaled",
                    "pvalue.scaled",
                    "rmsea.robust", "rmsea.ci.lower.robust", "rmsea.ci.upper.robust",
                    "cfi.robust","tli.robust","srmr")),3)

rownames     <- c("lower bounds", "two_factor", "three_factor")

fit_table    <- rbind(fit_criteria, fit_2factor)
fit_table    <- rbind(fit_table, fit_3factor)

colnames(fit_table)  <- columnnames
row.names(fit_table) <- rownames

pander(fit_table)
```

### Results

The correlation matrix helps assess the data.

```{r covariance_matrix}
# covariance matrix / "TK1a","TK567"
pander(create_matrix(CFA_TPACK_PST[,c("CK123", "CK5", "PCK2" , "PCK5" , "TK1b", "TK2", "TK34", "TCK1","TCK4", "TPK1","TPK4","TPACK2","TPACK4")])) # ,"TPK4"
```

To calculate the students TPACK-score we would only calculate the average of all the items, i.e. an integrative TPACK-model, where all the elements contributes to the students TPACK.

```{r scores}
TK    <- c("TK1b", "TK2", "TK34") # "TK1a","TK567"
PCK   <- c("CK123", "PCK2" , "PCK5" , "CK5")
TPACK <- c("TPK1","TPK4","TPACK2","TCK1","TPACK4", "TCK4") # , "TPK4"

TPACK_PST_items <- list("TK" = TK,
                        "PCK" = PCK,
                        "TPACK" = TPACK) # everyone contributes til TPACK; c(TK,"TCK4",TPK, "CK123","PCK2","PCK5"))

df    <- CFA_TPACK_PST[,unlist(TPACK_PST_items[1:3])]
describe(df)      # with missing (see n)? (na.rm=TRUE)

score_TPACK <- scoreItems(TPACK_PST_items,df)
describe(score_TPACK$scores)
```

@@

Three models are tested for reliability. The CFA-model (the estimated correlation matrix), the correlation matrix from the observed variables, and the observed variables.

```{r omega_mle4}
omega0  <- omegaFromSem(model_three_factor)
omega0

# get correlation matrix of observed variables
cor_mat <- lavInspect(model_three_factor, what="cor.ov")
omega(cor_mat, nfactors=3)

omega(CFA_TPACK_PST[,c("CK123", "CK5", "PCK2" , "PCK5" , "TCK1","TCK4", "TPK1","TPK4","TPACK2","TPACK4", "TK1b", "TK2", "TK34")]) # ,"TPK4"
```

@@

I test for discriminatory and convergent validity following @ronkko_updated_2022

```{r}
# ref.: https://link.springer.com/article/10.1007/s10490-023-09871-y
source("../../scripts/measureQ.R")
```

```{r measureQ, eval=T}
measureQ(model_three_factor, CFA_TPACK_PST)
```

@@

The results are discussed in @karlsen_assessing_2024.
